Typed Lua: An Optional Type System for Lua

### 2014-03-10 doc/paper/typedlua.tex
finished rewrite of section 2.3

Typed Lua lets assignment change the type of a local variable in
cases that respect a simple heuristic: the new type must be a subtype
of the previous one, and only applies for the remainder of the
current scope.

@@

Ideally, we want the type of {\tt person} to change as the
table gets built, from $\{\}$ to $\{$``firstname": $\String\}$
and finally to $\{$``firstname": $\String$, ``lastname":
 $\String\}$. This is tricker than the type change introduced
by assignment that we saw in Section 2.2, as what is changing
is not just the type of the variable {\tt person} but the
type of the value that {\tt person} points to.

>>>>

Typed Lua lets assignment change the type of a local variable in
cases that respect a simple heuristic: the new type must be a
subtype of the previous one, and the variable must be local
to the current function. The new type only applies for the
remainder of the current scope.

@@

Ideally, we want the type of {\tt person} to change as the
table gets built, from $\{\}$ to $\{$``firstname": $\String\}$
and finally to $\{$``firstname": $\String$, ``lastname":
 $\String\}$. This is tricker than the type change introduced
by assignment that we saw in Section 2.2, as what is changing
is not just the type of the variable {\tt person} but the
type of the value that {\tt person} points to. This is safe
in the example above, but not in the example below:

\begin{verbatim}
    local bogus = { firstname = 1 }
    local person: {} = bogus
    person.firstname = "Lou"
    person.lastname = "Reed"
\end{verbatim}

The assignment on line two is perfectly legal, as the type
of {\tt bogus} is a subtype of $\{\}$. But changing the type
of {\tt person} would be unsound: {\tt person.firstname} is
now a $\String$, but {\tt bogus.firstname} is still typed
as a $\Number$. We do not even need to declare a type for
aliasing to be a problem:

\begin{verbatim}
    local person = {}
    local bogus = person
    bogus.firstname = 1
    person.firstname = "Lou"
    person.lastname = "Reed"
\end{verbatim}

Taken individually, the changes to the type of the two
variables look ok, but aliasing makes one of them unsound.
The location of the change also matters, as the next example
shows:

\begin{verbatim}
    local person = {}
    local bogus = { firstname = 1 }
    do
      person.firstname = 1
      bogus = person
    end
    do
      person.firstname = "Lou"
    end
    -- bogus.firstname is now "Lou"
\end{verbatim}

The initial type of {\tt person} in all of these examples is
$\{\}$, but the {\em origin} of this type judgment matters
on whether it is sound to allow a change to the type of
{\tt person} or not, even if the change is always towards
a subtype of the current type.

Typed Lua tags a variable with a table type as either
{\em open} or {\em closed}. If a variable gets its type
from a table constructor then it is open, otherwise it is
always closed. The type of an open variable may change
by field assignment, subject to two restrictions: the
variable must be local to the current block, and the new type
must be a subtype of the old type.

Using a variable with an open type can also trigger a
type change, if the type of the missing fields has
{\tt nil} as a possible value. This lets the programmer
incrementally create an instance of an interface with
an optional type:

    local user = {}
    user.firstname = "Lou"
    user.lastname = "Reed"
    local person: Person = user
\end{verbatim}

Table types are the foundation for modules and objects in
Typed Lua. Type changes triggered by field assignment are
also an important part of Typed Lua's support for the
idiomatic definition of Lua modules, which are the
subject of the next section.
